---
title: "Salary Trends in Data Science "
author: "TEAM 3 - Jeongmin An, Yeobi Hobson, Sameer Batra, Aditi Shukla"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

# **Data Science Salary Trends (2020-2024)**  
#### *Exploring how salaries have evolved and the key factors influencing earnings*

---

## **Table of Contents**
- [Introduction](#introduction)
- [Loading Dataset](#.loading Dataset)
- [Data Cleaning](#.data cleaning)
- [Exploratory Data Analysis](#exploratory-data-analysis)
- [Statistical Analysis & Hypothesis Testing](#statistical-analysis-hypothesis-testing)
- [Conclusion](#conclusion)
- [References](#references)

---

## **1. Introduction**
Data science is a rapidly evolving field, and salaries vary widely based on experience, job role, work model, and company size.  
This analysis aims to uncover **salary trends from 2020 to 2024**, examining how different factors influence earnings.  

Our key research questions:  
- How have salaries changed over time?
- What is the relationship between experience and salary?
- Do work models (remote vs. in-person) impact earnings?
- How does company size influence salary distribution?

---



## **2. Loading Dataset**
**Source:** Kaggle Dataset (6,599 observations)  
**Years Covered:** 2020-2024  
```{r include=TRUE, echo=TRUE, results='markup'}
#load packages
library(dplyr)
library(ggplot2)

#load data
salaries = read.csv("data_science_salaries.csv")
```



## **3. Data cleaning**
### Check for missing values
```{r include=TRUE, echo=TRUE, results='markup'}
#check for missing values
total_na <- sum(is.na(salaries))
missing_summary <- colSums(is.na(salaries))
cat("Total missing values:", total_na, "\n")
print(missing_summary)

# Display first few rows
head(salaries)
```
Since no missing values are found, we proceed without handling missing data.

### Check data structure and summary
```{r include=TRUE, echo=TRUE, results='markup'}
str(salaries)
```

```{r include=TRUE, echo=TRUE, results='markup'}
summary(salaries)
```

### Check employee residence distribution
```{r include=TRUE, echo=TRUE, results='hide'}
# Create the grouped dataframe
country_counts <- as.data.frame(table(salaries$employee_residence))
colnames(country_counts) <- c("Country", "Frequency")

# Arrange by Frequency in ascending order
country_counts <- country_counts[order(country_counts$Frequency, decreasing = TRUE), ]

# View the result
country_counts
```

```{r include=TRUE, echo=TRUE, results='markup'}
# Create the barplot
ggplot(country_counts, aes(x = reorder(Country, -Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Frequency of Employee Residence by Country",
       x = "Country",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Filter for US-Based Employees & Remove unnecessary columns
```{r include=TRUE, echo=TRUE, results='markup'}
# Filter data(US)
us_residents_df <- subset(salaries, employee_residence == "United States")

##Since we are only working with us_residents dataframe we need to delete salary, salary_currency
us_residents_df <- subset(us_residents_df, select= -c(employee_residence, salary, salary_currency))

head(us_residents_df)
```

### Detecting Outliers Using IQR (Boxplot Method)
```{r include=TRUE, echo=TRUE, results='hide'}
# Function to find outliers using IQR and return entire rows
find_outliers <- function(data, column_name) {
  column <- data[[column_name]]
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  
  # Define lower and upper bounds
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  # Identify outliers (rows where the value is outside the bounds)
  outliers <- data[column < lower_bound | column > upper_bound, ]
  return(outliers)
}

# Apply function to salary and salary_in_usd
outliers_salary_usd <- find_outliers(us_residents_df, "salary_in_usd")
outliers_salary_usd

```

```{r include=TRUE, echo=TRUE, results='markup'}
# Boxplot for Salary in USD
ggplot(us_residents_df, aes(y = salary_in_usd)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red") +
  labs(title = "Boxplot of Salary in USD", y = "Salary in USD")
```

### Remove outliers
```{r include=TRUE, echo=TRUE, results='hide'}
# Get the indices of the rows with outliers in salary and salary_in_usd
outlier_rows_salary_usd <- which(us_residents_df$salary_in_usd %in% outliers_salary_usd$salary_in_usd)

# Remove outliers from the original dataset using subset
cleaned_salaries <- us_residents_df[-outlier_rows_salary_usd, ]
cleaned_salaries
```
```{r}
str(cleaned_salaries)
```
After removing the outliers, there are 5204 values in the dataframe. This means that 101 values were outliers.
```{r include=TRUE, echo=TRUE, results='markup'}
ggplot(cleaned_salaries, aes(y = salary_in_usd)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red") +
  labs(title = "Boxplot of Salary in usd", y = "Salary")
```

### Detecting Outliers Using the ezids package
```{r include=TRUE, echo=TRUE, results='markup'}

library(ezids)

# Outlier Removal + Visualization
us_residents_clean <- ezids::outlierKD2(us_residents_df, salary_in_usd, rm = TRUE, boxplt = TRUE, qqplt = TRUE)

# Check normality after removing outliers 
qqnorm(us_residents_clean$salary_in_usd, main = "QQ-Plot for Salary After Outlier Removal", col = "blue")
qqline(us_residents_clean$salary_in_usd, col = "red")

```




## **4. Exploratory Data Analysis**
### 1) How have salary trends in data science evolved from 2020 to 2024?
```{r include=TRUE, echo=TRUE, results='markup'}

#Check work_year value
unique(cleaned_salaries$work_year)

# Compute average salary by year
salary_trends <- cleaned_salaries %>%
  group_by(work_year) %>%
  summarise(
    Avg_Salary_USD = mean(salary_in_usd, na.rm = TRUE),
    Count = n()
  )

print(salary_trends)

# Line plot for Salary Trends
ggplot(salary_trends, aes(x = work_year, y = Avg_Salary_USD)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "red", size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Salary Trends in Data Science (2020-2024)", x = "Year", y = "Average Salary in USD") +
  theme_minimal()

# Box plot
ggplot(cleaned_salaries, aes(x = as.factor(work_year), y = salary_in_usd, fill = as.factor(work_year))) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Salary Distribution by Year",x = "Year",y = "Salary in USD") +
  theme_minimal()

# ANOVA test
anova_result <- aov(salary_in_usd ~ as.factor(work_year), data = cleaned_salaries)
summary(anova_result)


```
The ANOVA test reveals a highly significant effect of work year on salary (F = 6.716, p = 2.18e-05), indicating that salaries in data science have significantly changed over time.

```{r}
library(dplyr)
library(ggplot2)

# plot bar graph of salaries from 2020 to 2024 according to work_model
ggplot(cleaned_salaries, aes(x = work_year, y = salary_in_usd, fill = work_models)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Salary Distribution by Year and Work Model", x = "Year", y = "Salary in USD") +
  theme_minimal()
```

### 2) How do salaries vary across different experience levels in data science?
```{r include=TRUE, echo=TRUE, results='markup'}

#Check experience_level value
unique(cleaned_salaries$experience_level)

# Compute average salary and salary in USD by experience level
experience_salary <- cleaned_salaries %>%
  group_by(experience_level) %>%
  summarise(
    Avg_Salary_USD = mean(salary_in_usd, na.rm = TRUE),
    Count = n()
  )

# Display summary table
print(experience_salary)

library(scales)

cleaned_salaries$experience_level <- factor(
  cleaned_salaries$experience_level,
  levels = c("Entry-level", "Mid-level", "Senior-level", "Executive-level")
)

# Visualize with Boxplot
ggplot(cleaned_salaries, aes(x = experience_level, y = salary_in_usd, fill = experience_level)) +
  geom_boxplot() +
  labs(title = "Salary Distribution by Experience Level", x = "Experience Level", y = "Salary in USD") +
  scale_y_continuous(labels = scales::comma)+
  theme_minimal()

library(dplyr)
filtered_data <- cleaned_salaries %>% 
  filter(experience_level == "Entry-level" & salary_in_usd  )

# ANOVA test
anova_result <- aov(salary_in_usd ~ experience_level, data = cleaned_salaries)
summary(anova_result)
```
The ANOVA test shows a highly significant effect of experience level on salary (F = 249.2, p < 2e-16), indicating substantial salary differences across experience levels.


### 3) What is the average salary difference between work models data science jobs?
```{r include=TRUE, echo=TRUE, results='markup'}

#Check work_models value
unique(cleaned_salaries$work_models)

# Compute average salary by work model
work_model_salary <- cleaned_salaries %>%
  group_by(work_models) %>%
  summarise(
    Avg_Salary_USD = mean(salary_in_usd, na.rm = TRUE),
    Count = n() 
  )
print(work_model_salary)

# Boxplot of Salary by Work Model
ggplot(cleaned_salaries, aes(x = work_models, y = salary_in_usd, fill = work_models)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) + 
  labs(title = "Salary Comparison: Remote vs. In-Person", x = "Work Model", y = "Salary in USD")+
  theme_minimal()

# ANOVA test
anova_result <- aov(salary_in_usd ~ work_models, data = cleaned_salaries)
summary(anova_result)

```
The ANOVA test identifies a statistically significant difference in salaries across work models (F = 4.969, p = 0.00698), suggesting that at least one work model has a significantly different average salary."


### 4) Difference in salaries with respect to company size
```{r include=TRUE, echo=TRUE, results='markup'}

#Check company_size value
unique(cleaned_salaries$company_size)

salary_summary <- cleaned_salaries %>%
  group_by(company_size) %>%
  summarise(
    mean_salary = mean(salary_in_usd, na.rm = TRUE),
    median_salary = median(salary_in_usd, na.rm = TRUE),
    count = n()
  )

print(salary_summary)


#Box plot for Salary distribution by company size
ggplot(cleaned_salaries, aes(x = company_size, y = salary_in_usd, fill = company_size)) +
  geom_boxplot() +
  labs(title = "Salary Distribution by Company Size", x = "Company Size", y = "Salary (USD)") +
  scale_y_continuous(labels = comma)+
  theme_minimal()

# ANOVA test
anova_result <- aov(salary_in_usd ~ company_size, data = cleaned_salaries)
summary(anova_result)

```
The ANOVA test reveals a statistically significant effect of company size on salary (F = 8.18, p = 0.000284), indicating that salaries vary significantly across different company sizes. 


### 5) Which job title earned the most in 2024?
```{r include=TRUE, echo=TRUE, results='markup'}

salaries_2024 <- cleaned_salaries %>%
  filter(work_year == 2024)

# average salaries by job title
top_job <- salaries_2024 %>%
  group_by(job_title) %>%
  summarise(
    mean_salary = mean(salary_in_usd, na.rm = TRUE),
    max_salary = max(salary_in_usd, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(mean_salary))
print(top_job)

#top 10
top_10_jobs <- top_job %>% slice_head(n = 10)

ggplot(top_10_jobs, aes(x = reorder(job_title, mean_salary), y = mean_salary, fill = job_title)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(title = "Top 10 Highest Paying Job Titles in 2024", x = "Job Title", y = "Average Salary (USD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

highest_paid_job <- salaries_2024 %>%
  slice_max(salary_in_usd, n = 1)  
print(highest_paid_job)
```


### 5-1)  What are the most common job title (Which job title hire the most)?
#### Filter data by job title
```{r include=TRUE, echo=TRUE, results='markup'}
# Step 1: Calculate the frequencies of each job title
job_title_freq <- table(cleaned_salaries$job_title)

# Step 2: Create the frequency table as a data frame for plotting
job_title_freq_df <- as.data.frame(job_title_freq)
colnames(job_title_freq_df) <- c("job_title", "frequency") 

# Plotting the histogram
library(ggplot2)

ggplot(job_title_freq_df, aes(x = reorder(job_title, -frequency), y = frequency)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +  # Bar plot for categorical data
  labs(
    title = "Job Title Frequencies",  # Plot title
    x = "Job Title",  # X-axis label
    y = "Frequency"   # Y-axis label
  ) +
  theme_minimal() +  # Minimal theme
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability
```
Since this representation does not seem right we need to group the job_title categories that are less than 20 into other category and then convert the frequencies into logarthmic scale for better representation.

```{r include=TRUE, echo=TRUE, results='markup'}
freq_table <- table(cleaned_salaries$job_title)

freq_df <- as.data.frame(freq_table, stringsAsFactors = FALSE)

colnames(freq_df) <- c("job_title", "frequency")

threshold <- 20

freq_df$job_title[freq_df$frequency <= threshold] <- "Other"

freq_df <- aggregate(frequency ~ job_title, data = freq_df, sum)

freq_df <- freq_df[order(-freq_df$frequency), ]

cleaned_salaries$job_title <- ifelse(
  cleaned_salaries$job_title %in% freq_df$job_title,
  cleaned_salaries$job_title, 
  "Other"                     
)
#cleaned_salaries
```
Rather than keeping job titles that appeared less than 20 times as separate titles, we aggregated them into "other" keyword.

```{r include=TRUE, echo=TRUE, results='hide'}
freq_df
```

```{r include=TRUE, echo=TRUE, results='markup'}
ggplot(freq_df, aes(x = reorder(job_title, -frequency), y = frequency)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Bar plot
  labs(
    title = "Job Title Frequencies",  # Plot title
    x = "Job Title",  # X-axis label
    y = "Frequency"   # Y-axis label
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels
```
The above graph shows that job titles are right skewed.

```{r include=TRUE, echo=TRUE, results='hide'}
freq_df
```

```{r include=TRUE, echo=TRUE, results='hide'}
cleaned_salaries
unique(cleaned_salaries$employment_type)
```

```{r include=TRUE, echo=TRUE, results='markup'}
str(cleaned_salaries)
```



## **5. Statistical Analysis & Hypothesis Testing**
### Shapiro-Wilk test
```{r include=TRUE, echo=TRUE, results='markup'}
set.seed(123)
# Subset the data into Full-time and Part-time groups
full_time_data <- cleaned_salaries[cleaned_salaries$employment_type == "Full-time", ]
# using only 5000 because shapiro only accepts sample size between 3 to 5000
full_time_sample <- full_time_data[sample(nrow(full_time_data), 5000), ]
part_time_data <- cleaned_salaries[cleaned_salaries$employment_type == "Part-time", ]
contract_data <- cleaned_salaries[cleaned_salaries$employment_type=="Contract", ]

# Conduct the Shapiro-Wilk test for Full-time salaries
shapiro_test_full_time <- shapiro.test(full_time_sample$salary_in_usd)
print(shapiro_test_full_time)

# Conduct the Shapiro-Wilk test for Part-time salaries
shapiro_test_part_time <- shapiro.test(part_time_data$salary_in_usd)
print(shapiro_test_part_time)

shapiro.test(contract_data$salary_in_usd)
```

Null Hypothesis: The salary data for Full-time employees is normally distributed.
Alternate Hypothesis: The salary data for Full-time employees is not normally distributed.
Results: Since the p-value is less than 2.2e-16, which is far smaller than the significance level (α = 0.05), we reject the null hypothesis. This suggests that there is enough evidence to conclude that the salary data for Full-time employees is not normally distributed.

Null Hypothesis: The salary data for Part-time employees is normally distributed.
Alternate Hypothesis: Part-time salary data does not significantly deviate from normality, meaning it is more likely to be normally distributed.
Results: Since the p-value is 0.4491, which is greater than the significance level (α = 0.05), we fail to reject the null hypothesis. This suggests that there is insufficient evidence to conclude that the salary data for Part-time employees is not normally distributed.

Null Hypothesis: The salary data for Contract employees is normally distributed.
Alternate Hypothesis: The salary data for Contract employees is not normally distributed.
Results: Since the p-value is 0.1172, which is greater than the significance level (α = 0.05), we fail to reject the null hypothesis. This suggests that there is insufficient evidence to conclude that the salary data for Contract employees is not normally distributed.

```{r include=TRUE, echo=TRUE, results='markup'}
data_manager <- cleaned_salaries[cleaned_salaries$job_title == "Data Manager", ]
shapiro.test(data_manager$salary_in_usd)

research_analyst <- cleaned_salaries[cleaned_salaries$job_title == "Research Analyst", ]
shapiro.test(research_analyst$salary_in_usd)

data_engineer <- cleaned_salaries[cleaned_salaries$job_title == "Data Engineer", ]
shapiro.test(data_engineer$salary_in_usd)

data_scientist <- cleaned_salaries[cleaned_salaries$job_title == "Data Scientist", ]
shapiro.test(data_scientist$salary_in_usd)

data_analyst <- cleaned_salaries[cleaned_salaries$job_title == "Data Analyst", ]
shapiro.test(data_analyst$salary_in_usd)

analytics_engineer <- cleaned_salaries[cleaned_salaries$job_title == "Analytics Engineer", ]
shapiro.test(analytics_engineer$salary_in_usd)
```
Null Hypothesis: The salary data for Data Manager is normally distributed.
Alternate hypothesis: The salary data for Data Manager is not normally distributed.
results: Since the p-value (0.1905) is greater than the significance level (α = 0.05), we fail to reject the null hypothesis. This suggests that there is not enough evidence to conclude that the salary data for Data Manager is not normally distributed. Therefore, we can assume that the data follows a normal distribution.

Null Hypothesis: The salary data for research analyst is normally distributed.
Alternate hypothesis: The salary data for research analyst is not normally distributed.
results: Since the p-value (0.0003165) is much less than the significance level (α = 0.05), we reject the null hypothesis. This indicates that there is enough evidence to conclude that the salary data for Research Analysts is not normally distributed.

Final Result: From the above results we can conclude that only "Data Manager" and "Analytics Engineer" have normal distribution

```{r include=TRUE, echo=TRUE, results='markup'}
qqnorm(data_manager$salary_in_usd, main = "Q-Q Plot for Data Manager Salaries")
qqline(data_manager$salary_in_usd, col = "red", lwd = 2)

qqnorm(analytics_engineer$salary_in_usd, main = "Q-Q Plot for Analytics Engineer Salaries")
qqline(analytics_engineer$salary_in_usd, col = "red", lwd = 2)
```

### T-test
```{r include=TRUE, echo=TRUE, results='markup'}
t_test_result <- t.test(data_manager$salary_in_usd, analytics_engineer$salary_in_usd)

# Print the t-test result
print(t_test_result)
```
Null Hypothesis: The average salaries of Data Managers and Analytics Engineers are equal.
Alternative Hypothesis: The average salaries of Data Managers and Analytics Engineers are not equal.
Results: Given that the p-value is 1.513e-08 (which is much smaller than 0.05), we reject the null hypothesis. This means that there is a statistically significant difference in the average salaries of Data Managers and Analytics Engineers, with Analytics Engineers earning more on average than Data Managers.

```{r include=TRUE, echo=TRUE, results='markup'}
#Conduct shapiro test and t-test if applicable on work models column
unique(cleaned_salaries$work_models)
```


```{r include=TRUE, echo=TRUE, results='markup'}
remote <- cleaned_salaries[cleaned_salaries$work_models == "Remote", ]
shapiro.test(remote$salary_in_usd)

on_site_salaries <- cleaned_salaries[cleaned_salaries$work_models == "On-site", ]
shapiro.test(on_site_salaries$salary_in_usd)

hybrid <- cleaned_salaries[cleaned_salaries$work_models == "Hybrid", ]
shapiro.test(hybrid$salary_in_usd)
```
Null Hypothesis: All of the work models are normally distributed
Alternate Hypothesis: None of the work models are normally distributed
Results: Since all of the p-values(7.492e-12, 2.2e-16, 0.03666) are less than 0.05 we reject the null hypothesis. None of the work models (Remote, On-site, or Hybrid) have salaries that follow a normal distribution based on the Shapiro-Wilk test results.

```{r include=TRUE, echo=TRUE, results='hide'}
cleaned_salaries
```

### Chi-Squared Test
#### conducting chi-squared test between employment_type and work_models 
```{r include=TRUE, echo=TRUE, results='markup'}
table_data <- table(cleaned_salaries$employment_type, cleaned_salaries$work_models)
chisq_test <- chisq.test(table_data)
cat("\n\n")
cat("Chi-Squared Test for Employment Type and Work Models")
print(chisq_test)
chisq_test$expected
```
Null Hypothesis: There is no association between employment type and work model (they are independent).
Alternate Hypothesis: There is an association between employment type and work model (they are dependent).
Results: Since the p-value(6.673e-05) is less than 0.05, we reject the null hypothesis. There is a statistically significant association between employment type and work model. *However, we should be cautious about these results because, for a chi-squared test, the expected frequencies should be greater than 5. In this case, the expected counts for Hybrid: Contract, Contract:Remote, Part-time: Hybrid, Part-time: On-site, and Part-time: Remote are all less than 5, which may affect the reliability of the test.*

#### conducting chi-squared test between company_size and work_models 
```{r include=TRUE, echo=TRUE, results='markup'}
table_data <- table(cleaned_salaries$company_size, cleaned_salaries$work_models)
chisq_test <- chisq.test(table_data)
cat("\n\n")
cat("Chi-Squared Test for Company Size and Work Models")
print(chisq_test)
chisq_test$expected
```
Null Hypothesis: There is no association between company_size and work model (they are independent).
Alternate Hypothesis: There is an association between company_size and work model (they are dependent).
Results: Since the p-value(<2.2e-16) is less than 0.05, we reject the null hypothesis. There is a statistically significant association between company_size and work model. *However, we should be cautious about these results because, for a chi-squared test, the expected frequencies should be greater than 5. In this case, the expected counts for Hybrid: Large, small:hybrid are all less than 5, which may affect the reliability of the test.*

#### conducting chi-squared test between experience level and work_models
```{r include=TRUE, echo=TRUE, results='markup'}
table_data <- table(cleaned_salaries$work_models, cleaned_salaries$experience_level)
chisq_test <- chisq.test(table_data)
cat("\n\n")
cat("Chi-Squared Test for Experience Level and Work Models")
print(chisq_test)
chisq_test$expected
```
Null Hypothesis: There is no association between work model and experience level (they are independent).
Alternate Hypothesis: There is an association between work model and experience level (they are dependent).
Results: Since the p-value(<2.2e-16) is less than 0.05, we reject the null hypothesis. There is a statistically significant association between work model and experience level. *However, we should be cautious about these results because, for a chi-squared test, the expected frequencies should be greater than 5. In this case, the expected counts for Hybrid: Entry-level, and Hybrid:Executive-level are all less than 5, which may affect the reliability of the test.*

#### conducting chi-squared test between experience level and employement type
```{r include=TRUE, echo=TRUE, results='markup'}
table_data <- table(cleaned_salaries$experience_level, cleaned_salaries$employment_type)
chisq_test <- chisq.test(table_data)
cat("\n\n")
cat("Chi-Squared Test for Experience Level and Employment Type")
print(chisq_test)
chisq_test$expected
```
Null Hypothesis: There is no association between experience level and employment type(they are independent).
Alternate Hypothesis: There is an association between experience level and employment type(they are dependent).
Results: Since the p-value(3.076e-09) is less than 0.05, we reject the null hypothesis. There is a statistically significant association between experience level and employment type. *However, we should be cautious about these results because, for a chi-squared test, the expected frequencies should be greater than 5. In this case, the expected counts for Entry-level:contract, Executive-level:contract, Entry-level:Part-time, Executive-level and part-time, mid-level:contract, mid-level:part-time and senior-level:part-time are all less than 5, which may affect the reliability of the test.*




## **6. Conclusion**
📌 **Key Takeaways:**  
**Salaries have increased from 2020 to 2024**, likely due to growing demand.  
**Experience level significantly impacts salary**, with senior professionals earning much more.  
**Remote jobs offer higher salaries** than on-site roles.  
**Larger companies tend to pay better** than small organizations.  
**Statistical tests confirm significant relationships** between salary, experience, work model, and company size.



## **References**
1. Kaggle: Data Science Salaries Dataset  
2. Tidyverse Package Documentation  
3. R Graphics Cookbook


---

------------------------------------------------------------------------
# This section finalizes the dataset for model training and evaluation.

```{r}
str(cleaned_salaries)
```
```{r}
cleaned_salaries$job_title <- as.factor(cleaned_salaries$job_title)
cleaned_salaries$employment_type <- as.factor(cleaned_salaries$employment_type)
cleaned_salaries$work_models <- as.factor(cleaned_salaries$work_models)
cleaned_salaries$company_size <- as.factor(cleaned_salaries$company_size)

cleaned_salaries <- cleaned_salaries %>%
  select(-company_location)

str(cleaned_salaries)
```

```{r}
cleaned_salaries <- cleaned_salaries[order(cleaned_salaries$work_year), ]
train_data <- subset(cleaned_salaries, work_year < 2024)
test_data  <- subset(cleaned_salaries, work_year >= 2024)

X_train <- subset(train_data, select = -salary_in_usd)
X_train <- as.data.frame(X_train)
y_train <- train_data$salary_in_usd
y_train <- as.data.frame(y_train)

X_test <- subset(test_data, select = -salary_in_usd)
y_test <- test_data$salary_in_usd
```


#linear regression
```{r}
model_lm <- lm(train_data$salary_in_usd ~ ., data = train_data)
lm_preds <- predict(model_lm, newdata = X_test)
```

```{r}
lm_rmse <- sqrt(mean((lm_preds - y_test)^2))
lm_mae <- mean(abs(lm_preds - y_test))
lm_r2 <- summary(model_lm)$r.squared
lm_mape <- mean(abs((y_test - lm_preds) / y_test)) * 100

cat("Linear regression RMSE:", round(lm_rmse, 4), "\n")
cat("Linear regression MAE:", round(lm_mae, 4), "\n")
cat("Linear regression R2:", round(lm_r2, 4), "\n")

results_df <- data.frame(
  Metric = c("RMSE", "MAE", "R2", "MAPE"),
  Value = c(round(lm_rmse, 4), round(lm_mae, 4), round(lm_r2, 4), round(lm_mape, 4))
)
```
```{r}
summary(model_lm)
results_df
```

```{r}
min(cleaned_salaries$salary_in_usd)
max(cleaned_salaries$salary_in_usd)
```


#XGBoost
```{r}
# Load libraries
library(tidyverse)
library(caret)
library(xgboost)
library(ggplot2)

train_matrix <- train_data %>%
  mutate(across(c(job_title, employment_type, work_models, company_size, experience_level), as.integer)) %>%
  select(salary_in_usd, job_title, employment_type, work_models, company_size, experience_level, work_year)

test_matrix <- test_data %>%
  mutate(across(c(job_title, employment_type, work_models, company_size, experience_level), as.integer)) %>%
  select(salary_in_usd, job_title, employment_type, work_models, company_size, experience_level, work_year)

X_train <- train_matrix %>% select(-salary_in_usd)
y_train <- train_matrix$salary_in_usd

X_test <- test_matrix %>% select(-salary_in_usd)
y_test <- test_matrix$salary_in_usd

xgb_train <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
xgb_test <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)

xgb_model <- xgboost(
  data = xgb_train,
  objective = "reg:squarederror",
  nrounds = 100,
  verbose = 0
)

xgb_preds <- predict(xgb_model, xgb_test)

xgb_rmse <- sqrt(mean((xgb_preds - y_test)^2))
xgb_mae <- mean(abs(xgb_preds - y_test))
xgb_r2 <- 1 - (sum((y_test - xgb_preds)^2) / sum((y_test - mean(y_test))^2))
cat("XGBoost Regression RMSE:", round(xgb_rmse, 2), "\n")
cat("XGBoost Regression MAE:", round(xgb_mae, 2), "\n")
cat("XGBoost Regression R2:", round(xgb_r2, 2), "\n")

#xgb_preds <- xgb_preds * (salary_max - salary_min) + salary_min
#y_test <- y_test * (salary_max - salary_min) + salary_min
xgb_mape <- mean(abs((xgb_preds - y_test ) / y_test)) * 100

results_df <- data.frame(
  Metric = c("RMSE", "MAE", "R2", "MAPE"),
  Value = c(round(xgb_rmse, 4), round(xgb_mae, 4), round(xgb_r2, 4), round(xgb_mape, 4))
)

ggplot(data.frame(actual = y_test, predicted = xgb_preds),
       aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5, color = "purple") +
  geom_abline(color = "blue", linetype = "dashed") +
  labs(
    title = "Actual vs Predicted Salary (XGBoost)",
    x = "Actual Salary (Scaled)",
    y = "Predicted Salary (Scaled)"
  ) +
  theme_minimal()

results_df
```

#Training a random forest regression model

```{r}
library(randomForest)

set.seed(42)
rf_model <- randomForest(
  salary_in_usd ~ ., 
  data = train_data,
  ntree = 500,       
  importance = TRUE
)

rf_pred <- predict(rf_model, newdata = test_data)
```
Some predictions were much higher or lower than the actual results → The prediction error was quite large.


# ----------------------------
# Model Performance Evaluation
# ----------------------------
# 6) Calculating RMSE, R², MAE
```{r}
rmse_val <- RMSE(rf_pred, test_data$salary_in_usd)
r2_val <- R2(rf_pred, test_data$salary_in_usd)
mae_val <- MAE(rf_pred, test_data$salary_in_usd)
rf_mape <- mean(abs((y_test - rf_pred) / y_test)) * 100

cat("📈 Random Forest Model Performance\n")
cat("RMSE:", round(rmse_val, 2), "\n")
cat("R²:", round(r2_val, 4), "\n")
cat("MAE:", round(mae_val, 2), "\n")
cat("MAPE:", round(rf_mape, 2), "\n")
varImpPlot(rf_model, main = "Variable Importance (Random Forest)")
```
When we used a random forest regression model to predict the salaries of data scientists, we found that the model's prediction accuracy was somewhat limited.
The RMSE of the model is approximately $50,704, and the MAE is approximately $40,788, showing that there is a significant difference between the predicted value and the actual salary. In particular, R^2 value is 0.2334, which means that the input variables explain only about 23.3% of the salary variation. This suggests that the model does not contain enough information to explain the salary.


Variable Importance :
According to the variable importance plot from our Random Forest model, experience level and job group are the most influential factors in predicting salary. Variables such as work model, company size, and employment type had minimal impact, indicating that role and seniority level are far more critical in compensation trends than where or how the job is performed.
Therefore, it seems that the application of additional advanced modeling techniques will be necessary.


# Converting dataset to classfication
```{r}
classification_salary <- cleaned_salaries
```

Converting Salaries into classification ranges. To determine the range we are using ranges that U.S. Bureau of Labor Statistics (BLS) use.  
```{r}
str(classification_salary)
```
U.S. Bureau of Labor Statistics (BLS) ranges:
  - < $30,000 – Low income
  - $30,000–$49,999 – Lower-middle income
  - $50,000–$74,999 – Middle income
  - $75,000–$99,999 – Upper-middle income
  - $100,000+ – High income

```{r}
classification_salary$income_category <- cut(
  classification_salary$salary_in_usd,
  breaks = c(-Inf, 30000, 50000, 75000, 100000, Inf),
  labels = c("Low income", "Lower-middle income", "Middle income", "Upper-middle income", "High income"),
  right = FALSE
)
classification_salary$income_category <- as.factor(classification_salary$income_category)
table(classification_salary$income_category)

```

```{r}
classification_salary <- classification_salary[classification_salary$income_category != "Low income", ]

classification_salary$income_category <- droplevels(classification_salary$income_category)

table(classification_salary$income_category)
```

```{r, include=TRUE, echo=FALSE}
str(classification_salary)
```


```{r}
classification_salary <- classification_salary %>% select(-salary_in_usd)
str(classification_salary)
```


```{r}
train_data <- classification_salary[classification_salary$work_year < 2024, ]
test_data  <- classification_salary[classification_salary$work_year == 2024, ]
```


```{r}
library(nnet)
model_multi <- multinom(income_category ~ job_title + experience_level + employment_type + work_models + company_size + work_year,
                        data = train_data)
```


```{r}
library(caret)
predictions <- predict(model_multi, newdata = test_data)
confusionMatrix(data = as.factor(predictions), reference = test_data$income_category)
log_reg_acc <- confusionMatrix(data = as.factor(predictions), reference = test_data$income_category)$overall['Accuracy']
```

Trying Random Forest
```{r}
library(randomForest)
library(caret)

rf_model <- randomForest(
  income_category ~ job_title + experience_level + employment_type + work_models + company_size + work_year,
  data = train_data,
  ntree = 2000,
  importance = TRUE        
)
```

```{r}
rf_predictions <- predict(rf_model, newdata = test_data)
confusionMatrix(rf_predictions, test_data$income_category)

rf_acc <- confusionMatrix(rf_predictions, test_data$income_category)$overall['Accuracy']
rf_acc
```

Using XGBoost
```{r}
library(xgboost)
library(caret)
library(data.table)
```

```{r}
str(classification_salary)
```

```{r}
train_data$income_category <- as.factor(train_data$income_category)
test_data$income_category <- as.factor(test_data$income_category)

#train_data$income_category <- as.numeric(train_data$income_category) - 1  # To start from 0
#test_data$income_category <- as.numeric(test_data$income_category) - 1

```

```{r}
train_matrix <- model.matrix(income_category ~ job_title + experience_level + employment_type + work_models + company_size - 1, data = train_data)
test_matrix <- model.matrix(income_category ~ job_title + experience_level + employment_type + work_models + company_size - 1, data = test_data)

dim(train_matrix)  # Check how many features
```

```{r}
#set.seed(123)  # For reproducibility
#library(caret)
#library(xgboost)

# Make sure the target is a factor
#train_data$income_category <- as.factor(train_data$income_category)

#train_control <- trainControl(
#  method = "cv",
#  number = 5,
#  verboseIter = TRUE
#)

#xgb_grid <- expand.grid(
#  nrounds = c(150, 300, 500),
#  max_depth = c(6, 7, 15),
#  eta = c(0.01, 0.1, 0),
#  gamma = c(0),
#  colsample_bytree = 0.8,
#  min_child_weight = 1,
#  subsample = 0.8
#)

#set.seed(123)
#xgb_tuned <- train(
#  income_category ~ .,
#  data = train_data,
#  method = "xgbTree",
#  trControl = train_control,
#  tuneGrid = xgb_grid,
#  metric = "Accuracy"
#)

```
Hyperparameter tuning for XGBoost shows that the best parameters are:
- nrounds: 150
- max_depth: 6
- eta: 0.1
- gamma: 0
- colsample_bytree: 0.8
- min_child_weight: 1
- subsample: 0.8

```{r}
str(train_data)
```

```{r}
xgb_trained <- train(
  income_category ~ .,
  data = train_data,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    nrounds = 150,
    max_depth = 6,
    eta = 0.1,
    gamma = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample = 0.8
  )
)
```

```{r}
xgb_trained
```

```{r}
xgb_predictions <- predict(xgb_trained, newdata = test_data, type = "raw")
xgb_predictions <- as.factor(xgb_predictions)
```

```{r}
unique(test_data$income_category)
unique(xgb_predictions)
```

```{r include=TRUE, echo=FALSE, results='markup'}
confusionMatrix(xgb_predictions, as.factor(test_data$income_category))

xgb_accuracy <- mean(xgb_predictions == test_data$income_category)
print(paste("Accuracy:", xgb_accuracy))
```

```{r include=TRUE, echo=FALSE, results='markup'}
importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix)
```
```{r}
# Save the xgb_tuned model
#saveRDS(xgb_trained, "xgb_tuned_model.rds")
```

```{r}
# Install the e1071 package if not already installed
#install.packages("e1071")
#install.packages("caret")  # for evaluation
#install.packages("data.table")  # for data manipulation (optional)

# Load libraries
library(e1071)
library(caret)
library(data.table)
```

```{r}
train_data$income_category <- as.factor(train_data$income_category)
test_data$income_category <- as.factor(test_data$income_category)
```

```{r}
str(test_data)
```

```{r}
naive_bayes_model <- naiveBayes(income_category ~ job_title + experience_level + employment_type + 
                                work_models + company_size, data = train_data)

# print(naive_bayes_model)
```

```{r}
nb_predictions <- predict(naive_bayes_model, newdata = test_data)

nb_predictions <- factor(nb_predictions, levels = levels(test_data$income_category))
```

```{r}
confusionMatrix(nb_predictions, test_data$income_category)

nb_accuracy <- mean(nb_predictions == test_data$income_category)
print(paste("Accuracy:", nb_accuracy))
```

```{r}
svc_model <- svm(income_category ~ job_title + experience_level + employment_type + 
                 work_models + company_size, data = train_data, type = "C-classification", kernel = "radial")

svc_predictions <- predict(svc_model, newdata = test_data)
svc_predictions <- factor(svc_predictions, levels = levels(test_data$income_category))

unique(svc_predictions)

confusionMatrix(svc_predictions, test_data$income_category)

svc_accuracy <- mean(svc_predictions == test_data$income_category)
print(paste("Accuracy:", svc_accuracy))
```

```{r}
model_accuracies <- data.frame(
  Model = c("SVC", "Random Forest", "Logistic Regression", "XGBoost", "Naive Bayes"),
  Accuracy = c(svc_accuracy, rf_acc, log_reg_acc, xgb_accuracy, nb_accuracy)
)
model_accuracies <- model_accuracies[order(-model_accuracies$Accuracy), ]
model_accuracies

```

```{r}
ggplot(model_accuracies, aes(x = Model, y = Accuracy)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Accuracy, 6)), vjust = -0.5) +
  labs(title = "Model Accuracies", x = "Model", y = "Accuracy") +
  theme_minimal()
```

